Supplementary Information

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
small updates
remove unnecessary latex formatting (not done)

[Kristina J. Anderson-Teixeira](0000-0001-8461-9713)^1,2^*,
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Ross Alexander^5^,
Craig D. Allen^6^,
Raquel Alfaro-Sánchez^7^,
Tala Awada^8^,
Jennifer L. Baltzer^7^,
Patrick J. Baker^9^,
[Joseph D. Birch](0000-0001-8644-7345)^10^,
Sarayudh Bunyavejchewin^11^,
Paolo Cherubini^12,13^,
Stuart J. Davies^2^,
Cameron Dow^1,14^,
Ryan Helcoski^1^,
[Jakub Kašpar](0000-0003-1780-6310)^15^,
[James Lutz]( 0000-0002-2560-0710)^16^,
Ellis Q. Margolis^17^,
Justin Maxwell^18^,
Sean McMahon^2,19^,
Camille Piponiot^1,2,20^,
Sabrina E. Russo^8,21^,
[Pavel Šamonil](0000-0002-7722-8797)^15^,
Anastasia E. Sniderhan^7^,
[Alan J. Tepley](0000-0002-5701-9613)^1,22^,
[Ivana Vašíčková](0000-0002-6070-5956)^15^, 
Mart Vlam^23^,
Pieter A. Zuidema^23^

@@

21. Center for Plant Science Innovation, University of Nebraska – Lincoln, USA 68588-0660

22. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

23. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands

@@

site_details_appendix_title<- "Appendix S1. Site Details"
DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
traditional_comparison_title <- "Appendix S4. Methods for comparing our approach with traditional methods"
changing_climate_title <- "Appendix S5. Dealing with rapidly changing climate and tree growth"

@@

GLS_BCNM_legend = "Figure S35. Best GLS models including climate and $DBH$ for BCNM"
GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, $DBH$, and year for BCNM"
GLS_HKK_legend = "Figure S37. Best GLS models including climate and $DBH$ for HKK"
GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, $DBH$, and year for HKK"
GLS_SCBI_legend = "Figure S39. Best GLS models including climate and $DBH$ for SCBI"
GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, $DBH$, and year for SCBI"
GLS_LDW_legend = "Figure S41. Best GLS models including climate and $DBH$ for LDW"
GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, $DBH$, and year for LDW"
GLS_HF_legend = "Figure S43. Best GLS models including climate and $DBH$ for HF"
GLS_HF_year_legend = "Figure S44. Best GLS models including climate, $DBH$, and year for HF "
GLS_ZOF_legend = "Figure S45. Best GLS models including climate and $DBH$ for ZOF"
GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, $DBH$, and year for ZOF"
GLS_NIO_legend = "Figure S47. Best GLS models including climate and $DBH$ for NIO"
GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, $DBH$, and year for NIO"
GLS_LT_legend = "Figure S49. Best GLS models including climate and $DBH$ for LT"
GLS_LT_year_legend = "Figure S50. Best GLS models including climate, $DBH$, and year for LT"
GLS_CB_legend = "Figure S51. Best GLS models including climate and $DBH$ for CB"
GLS_CB_year_legend = "Figure S52. Best GLS models including climate, $DBH$, and year for CB"
GLS_SC_legend = "Figure S53. Best GLS models including climate and $DBH$ for SC"
GLS_SC_year_legend = "Figure S54. Best GLS models including climate, $DBH$, and year for SC"

@@

RW_interactions_all_legend = "Figure S56. All signficant climate - $DBH$ interactions with $RW$ as the response metric."
BAI_interactions_all_legend ="Figure S57. All signficant climate - $DBH$ interactions with $RW$ as the response metric."

decadal_plot_BCNM_legend = "Figure S58. Relationships between $DBH$ and $BAI$ by decade for BCNM."
decadal_plot_HKK_legend = "Figure S59. Relationships between $DBH$ and $BAI$ by decade for HKK."
decadal_plot_SCBI_legend = "Figure S60. Relationships between $DBH$ and $BAI$ by decade for SCBI."
decadal_plot_LDW_legend = "Figure S61. Relationships between $DBH$ and $BAI$ by decade for LDW."
decadal_plot_HF_legend = "Figure S62. Relationships between $DBH$ and $BAI$ by decade for HF."
decadal_plot_ZOF_legend = "Figure S63. Relationships between $DBH$ and $BAI$ by decade for ZOF."
decadal_plot_NIO_legend = "Figure S64. Relationships between $DBH$ and $BAI$ by decade for NIO."
decadal_plot_LT_legend = "Figure S65. Relationships between $DBH$ and $BAI$ by decade for LT."
decadal_plot_CB_legend = "Figure S66. Relationships between $DBH$ and $BAI$ by decade for CB."
decadal_plot_SC_legend = "Figure S67. Relationships between $DBH$ and $BAI$ by decade for SC."

@@

Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
However, conventional tree-ring analysis approaches cannot simultaneously account for the effects of climate, tree size, and time on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined primary climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, $DBH$, and time have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
Interactions between $DBH$ and climate were common (44% of cases tested). 
Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with $DBH$.
Accounting for $DBH$, growth rate varied directionally over time in most species, declining in 90%.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
As a rigorous analytical framework for statistically modeling tree growth responses to the primary climate drivers, $DBH$, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.

@@

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
Over timeframes of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided (e.g., by sampling at sites with characteristics that result in trees being more climate-limited), and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon. Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**

*(keep mostly the same):*
Tree diameter at breast height ($DBH$) scales predictably with numerous variables affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].

*(keep mostly the same, but shift emphasis more towards ecology):*
Characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors. 
Moreover, carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- is known to decline as individual trees and forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_agerelated_2011]. 
However, tree size and time are almost inextricably linked (because each cored individual increases in $DBH$ through time) and difficult to disentangle due to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.

@@

All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
From among the full set of original $RW$ measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis. 
We also excluded records with insufficient sample size or anomalous growth patterns, including
(1) species with < 7 cores,
(2) cores with < 30 years of record,
(3) contiguous portions of cores containing large outliers ($RW$ > mean plus 5 x SD of $RW$ for the entire core),
and (4) the final 20 years prior to death of trees cored dead. 
The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
From analyses including $DBH$ (see below), we further excluded
(1) trees for which we lacked data required to reconstruct $DBH$,
(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
and (3) poorly represented tails of the $DBH$ distribution, starting where reconstructed $DBH$ (see below) was represented by < 3 conspecific trees.
In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with $DBH$ (Table `r sampling_details_table`).

For each year in the tree-ring records, we reconstructed $DBH$, as detailed in Appendix `r DBH_reconstruction`. 
We used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).

Once $DBH$ had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass ($AGB$). 
Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records (Appendix `r climate_correction`). 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation ($PPT$); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration ($PET$) and precipitation day frequency ($PDF$).
For the one riparian site, NIO, we tested for an effect of stream flow ($SF$), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S. Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw). 
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeo_2020].

@@

Our analysis consisted of two main steps: (1) identification of the primary climate drivers , and (2) combining these climate drivers, $DBH$, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable ($RW$, $BAI$, or $\Delta AGB$). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals. 
The smoothing parameters were automaticaly selected by the *gam* function by generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
Here, for each permutation, `climate` specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables, we moved forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

@@

For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.

>>>>

[Kristina J. Anderson-Teixeira](0000-0001-8461-9713)^1,2^*,
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Ross Alexander^5^,
Craig D. Allen^6^,
Raquel Alfaro-Sánchez^7^,
Tala Awada^8^,
Jennifer L. Baltzer^7^,
Patrick J. Baker^9^,
[Joseph D. Birch](0000-0001-8644-7345)^10^,
Sarayudh Bunyavejchewin^11^,
Paolo Cherubini^12,13^,
Stuart J. Davies^2^,
Cameron Dow^1,14^,
Ryan Helcoski^1^,
[Jakub Kašpar](0000-0003-1780-6310)^15^,
[James Lutz]( 0000-0002-2560-0710)^16^,
Ellis Q. Margolis^17^,
Justin Maxwell^18^,
Sean McMahon^2,19^,
Camille Piponiot^1,2,20^,
Sabrina E. Russo^21,22^,
[Pavel Šamonil](0000-0002-7722-8797)^15^,
Anastasia E. Sniderhan^7^,
[Alan J. Tepley](0000-0002-5701-9613)^1,22^,
[Ivana Vašíčková](0000-0002-6070-5956)^15^, 
Mart Vlam^23^,
Pieter A. Zuidema^23^

@@

21. School of Biological Sciences, University of Nebraska-Lincoln, USA 68588

22. Center for Plant Science Innovation, University of Nebraska – Lincoln, USA 68588

23. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

24. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands

@@

site_details_appendix_title<- "Appendix S1. Site Details"
DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of DBH"
climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
traditional_comparison_title <- "Appendix S4. Methods for comparing our approach with traditional methods"
changing_climate_title <- "Appendix S5. Dealing with rapidly changing climate and tree growth"

@@

GLS_BCNM_legend = "Figure S35. Best GLS models including climate and DBH for BCNM"
GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, DBH, and year for BCNM"
GLS_HKK_legend = "Figure S37. Best GLS models including climate and DBH for HKK"
GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, DBH, and year for HKK"
GLS_SCBI_legend = "Figure S39. Best GLS models including climate and DBH for SCBI"
GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, DBH, and year for SCBI"
GLS_LDW_legend = "Figure S41. Best GLS models including climate and DBH for LDW"
GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, DBH, and year for LDW"
GLS_HF_legend = "Figure S43. Best GLS models including climate and DBH for HF"
GLS_HF_year_legend = "Figure S44. Best GLS models including climate, DBH, and year for HF "
GLS_ZOF_legend = "Figure S45. Best GLS models including climate and DBH for ZOF"
GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, DBH, and year for ZOF"
GLS_NIO_legend = "Figure S47. Best GLS models including climate and DBH for NIO"
GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, DBH, and year for NIO"
GLS_LT_legend = "Figure S49. Best GLS models including climate and DBH for LT"
GLS_LT_year_legend = "Figure S50. Best GLS models including climate, DBH, and year for LT"
GLS_CB_legend = "Figure S51. Best GLS models including climate and DBH for CB"
GLS_CB_year_legend = "Figure S52. Best GLS models including climate, DBH, and year for CB"
GLS_SC_legend = "Figure S53. Best GLS models including climate and DBH for SC"
GLS_SC_year_legend = "Figure S54. Best GLS models including climate, DBH, and year for SC"

@@

RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with $RW$ as the response metric."
BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with $RW$ as the response metric."

decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and $BAI$ by decade for BCNM."
decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and $BAI$ by decade for HKK."
decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and $BAI$ by decade for SCBI."
decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and $BAI$ by decade for LDW."
decadal_plot_HF_legend = "Figure S62. Relationships between DBH and $BAI$ by decade for HF."
decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and $BAI$ by decade for ZOF."
decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and $BAI$ by decade for NIO."
decadal_plot_LT_legend = "Figure S65. Relationships between DBH and $BAI$ by decade for LT."
decadal_plot_CB_legend = "Figure S66. Relationships between DBH and $BAI$ by decade for CB."
decadal_plot_SC_legend = "Figure S67. Relationships between DBH and $BAI$ by decade for SC."

@@

Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
However, conventional tree-ring analysis approaches cannot simultaneously account for the effects of climate, tree size, and time on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined primary climate drivers, reconstructed tree diameter (DBH), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, DBH, and time have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
Interactions between DBH and climate were common (44% of cases tested). 
Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with DBH.
Accounting for DBH, growth rate varied directionally over time in most species, declining in 90%.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
As a rigorous analytical framework for statistically modeling tree growth responses to the primary climate drivers, DBH, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.

@@

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
Over timeframes of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided (e.g., by sampling at sites with characteristics that result in trees being more climate-limited), and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon. Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**

*(keep mostly the same):*
Tree diameter at breast height (DBH) scales predictably with numerous variables affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].

*(keep mostly the same, but shift emphasis more towards ecology):*
Characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors. 
Moreover, carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- is known to decline as individual trees and forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_agerelated_2011]. 
However, tree size and time are almost inextricably linked (because each cored individual increases in DBH through time) and difficult to disentangle due to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling DBH from time and detecting directional growth trends.

@@

All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
From among the full set of original $RW$ measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis. 
We also excluded records with insufficient sample size or anomalous growth patterns, including
(1) species with < 7 cores,
(2) cores with < 30 years of record,
(3) contiguous portions of cores containing large outliers ($RW$ > mean plus 5 x SD of $RW$ for the entire core),
and (4) the final 20 years prior to death of trees cored dead. 
The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
From analyses including DBH (see below), we further excluded
(1) trees for which we lacked data required to reconstruct DBH,
(2) trees for which there was a significant inconsistency between measured DBH and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) was represented by < 3 conspecific trees.
In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).

For each year in the tree-ring records, we reconstructed DBH, as detailed in Appendix `r DBH_reconstruction`. 
We used allometric equations between DBH and bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).

Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass ($AGB$). 
Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records (Appendix `r climate_correction`). 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation (PPT); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration (PET) and precipitation day frequency (PDF).
For the one riparian site, NIO, we tested for an effect of stream flow (SF), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S. Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw). 
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeo_2020].

@@

Our analysis consisted of two main steps: (1) identification of the primary climate drivers , and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable ($RW$, $BAI$, or $\Delta AGB$). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals. 
The smoothing parameters were automaticaly selected by the *gam* function by generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
Here, for each permutation, `climate` specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables, we moved forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

@@

For models including interactive effects of climate and DBH, we tested for interactions between first-order linear terms for climate variables and DBH.
